### 2016-09-08 analysis-note-v2/RTPC.tex
update the answers and minor text cleanings in the note
Clean up of the note RTPC chapter

~~~~At 6 GeV incident electron beam energy, the recoil $^{4}He$ nuclei from the coherent DVCS channel, have an average momentum around 200 MeV/c. CLAS cannot detect such low energy particles, for exemple protons are detected with a threshold of about 250 MeV/c, which will be even worse for helium. In order to detect such low-energy recoils a Radial TPC (RTPC), similar to the one used during the BoNuS run \cite{BONUS}, was added to CLAS. In 2009, the CLAS-EG6 group \cite{proposal} has built this RTPC, figure \ref{fig:RTPC2} shows a close photo of the RTPC, while figure \ref{fig:RTPC_xscetion} shows a schematic drawing of its internal structure. It is a 200 mm-long and 150 mm-diameter TPC, composed of two electronically separated modules, each covering about 150$^{\circ}$ of the azimuthal angle. The RTPC tracks the charged particles through the ionization centers they leave in the gas-filled detection volume. Then, the ionized electrons drift under the effect of an electric field and are amplified by a gas electron multiplier system before to be detected on collection pads.

@@

The RTPC electron collection system has 3200 readout pads. Each module of the 
RTPC has 40 rows and 40 columns of readout pads, each is 5 mm long and 4.45 mm 
wide. Each group of 16 pads is connected to a pre-amplifier before the recorded 
signals are carried to the acquisition electronics. There are 20 rows and 5 
columns of pre-amplifiers per module. This readout system records the charge 
information in time bins, in which the charge is measured in 
Analog-to-Digital-Converter (ADC) units, while the time is recorded in 
Time-to-Digital-Converter (TDC) units, in which each TDC is equal to 114 ns.  
This measurement indicating the time taken by the electrons to drift from the 
ionization point to the readout board.

@@

\section{RTPC calibration}
~~~~Reconstructing a trajectory from the recorded time information of the 
electrons requires a good knowledge of their drift speed and drift paths.  
Also, the gains of the readout pads are required to calculate $\frac{dE}{dx}$ 
from the recorded ADCs. The drift paths and the gain calibration of the RTPC 
require well identified events, so we decided to use elastic scattering 
($e^{4}He \rightarrow e^{4}He$). We had specific calibration runs with 1-pass 
beam energies of 1.024 and 1.269 GeV in order to enhance the cross section of 
the elastic process at large angles, which is highly suppressed at higher beam 
energy.\\

A GEANT4 simulation for the RTPC has been developed to help reproduce these 
events in order to properly extract the drift paths and the gains. Their 
extractions are based on comparing the experimentally identified $^4He$ elastic 
tracks to the GEANT4 simulated ones. Indeed, the kinematics of each elastic 
$^4He$ can be calculated from the electron measured in CLAS and then simulated 
in GEANT4. The output of the simulation is then compared with the measured 
signals to determine the drift paths. \\

This process is iterative since good tracks need to be identified at the 
beginning of the process. For this analysis, we started from a MAGBOLTZ 
calibration, which was then improved with the method described above with 
several iterations. In particular, changes in the alignment of the RTPC with 
respect to the beamline, detailed in CLAS-NOTE~\cite{eg6beamoffset}. In the 
following subsections, we present here only the selection of the RTPC tracks 
and elastic events for the last iteration of our calibration process. It is 
then followed by details on the techniques of calibration.

@@

\begin{figure}[tbp]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.38]{fig_rtpc/RTPC_track.png}
\vspace{-0.1in}
\caption{Schematic of a segment of the RTPC in a plane perpendicular to the 
beam line, with a $^4He$ track (in green) crossing the volume. See the text for 
description of the various elements and notations. $sdist$ and $edist$ are 
defined position as shown in this sketch.} \label{fig:RTPC_track}
\end{figure}

@@

\item We apply a cut on the vertex position in $z$ of the reconstructed tracks 
   to insure it is in the RTPC volume, i.e. $z$ $\in$ [-80,80] mm, as 
   illustrated in figure \ref{fig:rtpc_zz}.

\item We request a positive radius of curvature ($r_{0}>0$): the $^4He$ is a positively 
   charged nucleus and the reconstructed track in the RTPC must have positive 
   radius of curvature. In other words, it travels in a clockwise direction if 
   one looks into the electron beam. The curvature distribution of the 
   collected RTPC tracks is shown in figure \ref{fig:rtpc_ro}.

\item We check the quality of the helix fit with a $\chi^{2}$ cut (< 3.5). The 
   quality of the fit ($\chi^{2}$) is defined as:

\begin{equation}
   \chi^{2} = \frac{\displaystyle \left(\frac{DOCA}{\sigma_r}\right)^2 
       + \sum_{i = 1}^{ N_{pts}} \left(\frac{r^{pt}_{i} 
      - r^{helix}_{i} }{\sigma_{r}}\right)^{2}  + \left( \frac{\phi^{pt}_{i} - 
      \phi^{helix}_{i} }{\sigma_{\phi}}\right)^{2} + \left( \frac{z^{pt}_{i} - 
   z^{helix}_{i} }{\sigma_{z} } \right)^{2}}{N_{pts} - 4}
\end{equation}
where $\frac{DOCA}{\sigma_r}$ is the beam-spot constraint, $r^{pt}_{i}$, 
$\phi^{pt}_{i}$, $z^{pt}_{i}$ are the radial, azimuthal, longitudinal 
coordinates of each reconstructed hit "$i$". ($r^{helix}_{i}$,
$\phi^{helix}_{i}$, $z^{helix}_{i}$) is the point on the helix fit that is 
closest, in Euclidean (3D) space, to the reconstructed hit $i$. The resolutions 
are defined as ($\sigma_{r}$, $\sigma_{\phi}$, $\sigma_{z}$) = (0.53 mm, 
2$^{\circ}$, 1.2 mm) based on the size
of the readout pads were kept identical for all calibration iterations. These 
resolutions were derived by CLAS-BoNus experiment \cite{BONUS}. $N_{pts}$ is 
the total number of the hits in the chain.  Figure \ref{fig:rrtpc_X2} shows the 
$\chi^{2}$ distributions for the positive tracks which originate within the 
RTPC.

@@

\item $sdist$ ($\in$ [-2.0,2.0] mm) and $edist$ ($\in$ [-1,5] mm ) cuts: the 
   cuts are applied to ensure that a track is on time, i.e. the first 
   ionization point is close to the cathode and the last point is close to the 
   anode. For both $sdist$ and $edist$, positive values correspond to tracking 
   points inside the drift region, while negatives are outside the drift 
   region. So, in terms of the path length of the track starting from the 
   vertex, $sdist$ and $edist$ have opposite signs. The $sdist$ and $edist$ 
   distributions are shown in figures \ref{fig:rrtpc_sdist} and 
   \ref{fig:rrtpc_edist} respectively\footnote{The long tail on the positive 
      side of the $edist$ distribution is mostly due to electronic noise.  Note 
      that after applying all the elastic He-4 selection cuts, we observe that 
   most of the tails for both $edist$ and $sdist$ are removed.}.

@@

\subsubsection{Elastic selection}
   The elastic process on $^{4}He$ is defined as:
\begin{equation}
\centering
e(k) + ^{4}He(p) \rightarrow e(k') + ^{4}He(p')
\end{equation}
with the symbols in parenthesis representing the four-momenta of the particles. \\

 In addition to the previously good-track requirements, in order to select the elastic events we impose further constraints. The co-planarity between the scattered electron and the recoil $^4He$ is ensured using a $\Delta \phi$ (=$\phi_{e}-\phi_{^{4}He}$) cut as shown in Figure \ref{fig:delta_phi_elastic}. Like for $\Delta z$, the $\Delta \phi$ 2$\sigma$ cut is dependent on $z$ to account for variations. Our final parametrizations can be found in Appendix \ref{app:RTPC_appendix} as well.\\

@@

Constructing the invariant mass is one way to verify the efficiency of the previously described selection procedures. For a helium target, the invariant mass (W) is:\\
\begin{equation}
 W = \sqrt{-2E_{b}p_{e^{'}}(1-cos(\theta_{e^{'}})) + M_{^{4}He}^2 + 2(E_{b}-p_{e^{'}}) M_{^{4}He}}.
\end{equation}
with the conventions for the variables as in equation \ref{equ:theta_helium}.  
Figure \ref{fig:W_elastic} shows the W distribution of the events that have a 
good track in the RTPC and the identified elastic events. One can see the good 
agreement between the helium-4 real mass and the mean value of the identified 
elastic events. The difference in width is related to the different resolutions 
of the corresponding CLAS sectors. The reaction products are emitted back to 
back resulting in only small regions of overlapping acceptances between the 
RTPC and CLAS enhancing this effect. The reason for this is that local 
variations of the resolution can be much more important than ones averaged over 
the full detector.
\\

@@

\subsubsection{Checking the initial good track requirements}
~~~~In this section, we recheck the individual good track requirements 
presented in section \ref{good_track_req}, where we apply the other good track 
and elastic requirements and watch the individual variables.

Figure \ref{fig:2-z-vertex} presents the z-vertex distributions of the elastic 
events, where all the good track and elastic requirements were applied. One 
sees that the initial [-80:80] mm cuts on the helium z-vertex appear to be 
suitable cuts to ensure that the reconstructed tracks are within the physical 
volume of the RTPC.
\begin{figure}[tp]
  \centering
   \includegraphics[scale=0.34]{fig_rtpc/updates/elastic_z_vertex.png}
\caption{z-vertex distributions for the electrons of the identified elastic 
events in the two modules of the RTPC.}
\label{fig:2-z-vertex}
\end{figure}

In figure \ref{fig:2-r0}, we show the distribution of the radius of curvature 
for the identified elastic tracks in the RTPC. One sees that applying the other 
good track and elastic conditions will eliminate all the negative tracks and 
the tracks with $r_{0}$ less than 30 mm.
\begin{figure}[tp]
   \centering
\includegraphics[scale=0.34]{fig_rtpc/updates/r0_elastic_1p2GeV.png}
\caption{The radius of curvature distributions of the identified elastic 
$^{4}$He tracks in the two modules of the RTPC after applying the other good 
track and elastic events}.
\label{fig:2-r0}
\end{figure}

The distributions of $sdist$ and $edist$ for the identified elastic $^4$He 
elastic tracks, figure \ref{fig:2-sdist-edist}, show that most of the tails, 
which are mostly due to electronic noise, are removed. Therefore, making 
tighter cuts would not cause changes in the calibration as the tracks in the 
tail regions are cleaned by the other good track and elastic cuts.

@@

\subsection{drift paths and drift speed calibration}

@@

\section{RTPC tracking offsets}
TBC

>>>>

~~~~At 6 GeV incident electron beam energy, the recoil $^{4}He$ nuclei from the coherent DVCS channel, have an average momentum around 200 MeV/c. CLAS cannot detect such low energy particles, for exemple protons are detected with a threshold of about 250 MeV/c, which will be even higher for helium. In order to detect such low-energy recoils a Radial TPC (RTPC), similar to the one used during the BoNuS run \cite{BONUS}, was added to CLAS. In 2009, the CLAS-EG6 group \cite{proposal} has built this RTPC, figure \ref{fig:RTPC2} shows a close photo of the RTPC, while figure \ref{fig:RTPC_xscetion} shows a schematic drawing of its internal structure. It is a 200 mm-long and 150 mm-diameter TPC, composed of two electronically separated modules, each covering about 150$^{\circ}$ of the azimuthal angle. The RTPC tracks the charged particles through the ionization centers they leave in the gas-filled detection volume. Then, the ionized electrons drift under the effect of an electric field and are amplified by a gas electron multiplier system before to be detected on collection pads.

@@

The RTPC electron collection system has 3200 readout pads. Each module of the 
RTPC has 40 rows and 40 columns of readout pads, each is 5 mm long and 4.45 mm 
wide. Each group of 16 pads is connected to a pre-amplifier before the recorded 
signals are carried to the acquisition electronics. There are 20 rows and 5 
columns of pre-amplifiers per module. This readout system records the charge 
information in time bins, in which the charge is measured in 
Analog-to-Digital-Converter (ADC) units, while the 114 ns width of the time bins 
will be refered as Time-to-Digital-Converter (TDC) units.  
From this measurement, we deduce the time taken by the electrons to drift from the 
ionization point to the readout board.

@@

\section{RTPC calibration}
~~~~Reconstructing a trajectory from the recorded time information of the 
electrons requires a good knowledge of their drift speed and drift paths.  
Also, the gains of the readout pads are required to calculate $\frac{dE}{dx}$ 
from the recorded ADCs. The determination of the drift paths and the gain 
calibration of the RTPC 
require well identified events, so we decided to use elastic scattering 
($e^{4}He \rightarrow e^{4}He$). However, the cross section of 
the elastic process at detectable angles is highly suppressed at higher beam 
energy, so we had specific calibration runs with 1-pass 
beam energies of 1.024 and 1.269 GeV. \\

A GEANT4 simulation for the RTPC has been developed to help reproduce these 
events in order to properly extract the drift paths and the gains. Their 
extractions are based on comparing the experimentally identified $^4He$ elastic 
tracks to the GEANT4 simulated ones. Indeed, the kinematics of each elastic 
$^4He$ can be calculated from the electron measured in CLAS and then simulated 
in GEANT4. The output of the simulation gives an expected trajectory that we then
match to to the measured signals to determine the drift paths. \\

This process is iterative since good tracks need to be identified at the 
beginning of the process and as the calibration improves, more good tracks 
are identified. For this analysis, we started from a MAGBOLTZ 
calibration, which was then improved with the method described above with 
several iterations. In particular, it is important that the alignment of the 
RTPC with respect to the beamline is properly implemented (see details in
CLAS-NOTE~\cite{eg6beamoffset}). As the process evolved and was improved over 
time, we present in the following subsections only the method used for the last 
iteration of our calibration process.

@@

\begin{figure}[tbp]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.38]{fig_rtpc/RTPC_track.png}
\vspace{-0.1in}
\caption{Schematic of a segment of the RTPC in a plane perpendicular to the 
beam line, with a $^4He$ track (in green) crossing the volume. See the text for 
description of the various elements and notations. $sdist$ and $edist$ are 
distances between first and last hits and boundaries of the RTPC as shown in 
this sketch.} \label{fig:RTPC_track}
\end{figure}

@@

\item The vertex position in $z$ is in the RTPC volume, i.e. $z$ $\in$ [-80,80] mm, as 
   illustrated in figure \ref{fig:rtpc_zz}.

\item The radius of curvature is positive ($r_{0}>0$): the $^4He$ is a positively 
   charged nucleus and the reconstructed track in the RTPC must have positive 
   radius of curvature. In other words, it travels in a clockwise direction if 
   one looks into the electron beam. The curvature distribution of the 
   collected RTPC tracks is shown in figure \ref{fig:rtpc_ro}.

\item The helix fit is well reconstructed with a $\chi^{2} < 3.5$. The 
   quality of the fit ($\chi^{2}$) being defined as:
\begin{equation}
   \chi^{2} = \frac{\displaystyle \left(\frac{DOCA}{\sigma_r}\right)^2 
       + \sum_{i = 1}^{ N_{pts}} \left(\frac{r^{pt}_{i} 
      - r^{helix}_{i} }{\sigma_{r}}\right)^{2}  + \left( \frac{\phi^{pt}_{i} - 
      \phi^{helix}_{i} }{\sigma_{\phi}}\right)^{2} + \left( \frac{z^{pt}_{i} - 
   z^{helix}_{i} }{\sigma_{z} } \right)^{2}}{N_{pts} - 4}
\end{equation}
where $\frac{DOCA}{\sigma_r}$ is the beam-spot constraint, $r^{pt}_{i}$, 
$\phi^{pt}_{i}$, $z^{pt}_{i}$ are the radial, azimuthal, longitudinal 
coordinates of each reconstructed hit "$i$". ($r^{helix}_{i}$,
$\phi^{helix}_{i}$, $z^{helix}_{i}$) is the point on the helix fit that is 
closest to the reconstructed hit $i$. The resolutions 
are defined as ($\sigma_{r}$, $\sigma_{\phi}$, $\sigma_{z}$) = (0.53 mm, 
2$^{\circ}$, 1.2 mm) based only on the size
of the readout pads and were kept identical for all calibration iterations. These 
resolutions were derived by CLAS-BoNus experiment \cite{BONUS}. $N_{pts}$ is 
the total number of the hits in the chain. Figure \ref{fig:rrtpc_X2} shows the 
$\chi^{2}$ distributions for the positive tracks which originate within the 
RTPC.

@@

\item $sdist$ ($\in$ [-2.0,2.0] mm) and $edist$ ($\in$ [-1,5] mm ): these 
   cuts are applied to ensure that a track is on time, i.e. the first 
   ionization point is close to the cathode and the last point is close to the 
   anode. For both $sdist$ and $edist$, positive values correspond to tracking 
   points inside the drift region, while negatives are outside the drift 
   region. So, in terms of the path length of the track starting from the 
   vertex, $sdist$ and $edist$ have opposite signs. The $sdist$ and $edist$ 
   distributions are shown in figures \ref{fig:rrtpc_sdist} and 
   \ref{fig:rrtpc_edist} respectively\footnote{The long tail on the positive 
      side of the $edist$ distribution is mostly due to electronic noise.  Note 
      that after applying all the elastic He-4 selection cuts, we observe that 
   most of the tails for both $edist$ and $sdist$ are removed.}.

@@

\subsubsection{Elastic scattering event selection}
   The elastic process on $^{4}He$ is defined as:
\begin{equation}
\centering
e(k) + ^{4}He(p) \rightarrow e(k') + ^{4}He(p')
\end{equation}
with the symbols in parenthesis representing the four-momenta of the particles. \\

 In addition to the good-track requirements described above, to select elastic events we impose further constraints. The co-planarity between the scattered electron and the recoil $^4He$ is ensured using a $\Delta \phi$ (=$\phi_{e}-\phi_{^{4}He}$) cut as shown in Figure \ref{fig:delta_phi_elastic}. Like for $\Delta z$, the $\Delta \phi$ 2$\sigma$ cut is dependent on $z$ to account for systematic variations. Our final parametrizations can be found in Appendix \ref{app:RTPC_appendix} as well.\\

@@

Constructing the invariant mass is one way to verify the efficiency of the previously described selection procedures. For a helium target, the invariant mass (W) is:\\
\begin{equation}
 W = \sqrt{-2E_{b}p_{e^{'}}(1-cos(\theta_{e^{'}})) + M_{^{4}He}^2 + 2(E_{b}-p_{e^{'}}) M_{^{4}He}}.
\end{equation}
with the conventions for the variables as in equation \ref{equ:theta_helium}.  
Figure \ref{fig:W_elastic} shows the W distribution of the events that have a 
good track in the RTPC and the identified elastic events. One can see the good 
agreement between the helium-4 real mass and the mean value of the identified 
elastic events. The difference in width is related to differences in 
resolution in CLAS.
The large effect is due to the reaction products being emitted back to 
back resulting in a small overall acceptance for the process. This result is
therefore very sensitive to the local 
variations of the resolution, which can be rather important in CLAS.\\

@@

\subsubsection{Checking the initial good track requirements}
~~~~In this section, we study some of the good track requirements 
presented in section \ref{good_track_req}. We apply here all the good track 
and elastic requirements but the one related to the ploted variable.

Figure \ref{fig:2-z-vertex} presents the z-vertex distributions of the elastic 
events. One 
sees that the initial [-80:80] mm cuts on the helium z-vertex appear to be 
suitable cuts to ensure that the reconstructed tracks are within the physical 
volume of the RTPC.
\begin{figure}[tp]
  \centering
   \includegraphics[scale=0.34]{fig_rtpc/updates/elastic_z_vertex.png}
\caption{z-vertex distributions for the electrons of the identified elastic 
events in the two modules of the RTPC.}
\label{fig:2-z-vertex}
\end{figure}

In figure \ref{fig:2-r0}, we show the distribution of the radius of curvature 
for the identified elastic tracks in the RTPC. One sees that applying the other 
good track and elastic conditions eliminate all the negative tracks as well as
the tracks with $r_{0}$ less than 30 mm.
\begin{figure}[tp]
   \centering
\includegraphics[scale=0.34]{fig_rtpc/updates/r0_elastic_1p2GeV.png}
\caption{The radius of curvature distributions of the identified elastic 
$^{4}$He tracks in the two modules of the RTPC after applying the other good 
track and elastic events}.
\label{fig:2-r0}
\end{figure}

The distributions of $sdist$ and $edist$ for the identified elastic $^4$He 
elastic tracks, figure \ref{fig:2-sdist-edist}, show that the tails, 
which are mostly due to electronic noise, are removed. Therefore, making 
tighter cuts would not cause significant changes in the calibration as the 
tail regions are cleaned by the other good track and elastic cuts.

@@

\subsection{Drift paths and drift speed calibration}

@@
